---
title: "MLB Franchise Management Efficiency"
author: "David Lucey"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 7
    theme: flatly
vignette: >
  %\VignetteIndexEntry{MLB Franchise Management Efficiency}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center"
)
```

## Background: What is WAR?

**Wins Above Replacement (WAR)** measures how many more wins a player contributed
than a freely available minor-league call-up would have.  A WAR of 5 means the
player was worth 5 extra wins to his team.  FanGraphs publishes separate WAR for
batters (fWAR: hitting + baserunning + fielding) and pitchers (FIP-based).  We
use FanGraphs WAR here because it extends back to 1985 and is available via the
`baseballr` package.

| WAR range | Interpretation |
|----------:|----------------|
| < 0 | Below replacement level |
| 0–2 | Bench / back-of-rotation player |
| 2–5 | Solid regular or mid-rotation starter |
| 5–8 | All-Star calibre |
| > 8 | MVP / Cy Young calibre |

## The Question We Are Answering

Every team can buy payroll.  The question is: **who converts dollars into wins,
playoff appearances, and ultimately World Series rings — and why do some
franchises consistently fail to do so despite spending heavily?**

We score each franchise across five management dimensions using percentile ranks:

1. **FA efficiency** — median $/WAR paid to veteran free agents
2. **Dead money** — salary committed to players who produced zero or negative WAR
3. **Homegrown pipeline** — share of WAR from players developed internally
4. **October WAR retention** — how much of a team's regular-season talent showed
   up in the postseason (proxy for health and roster depth)
5. **Playoff achievement** — how deep teams went when they made the postseason

## Data Coverage

- **Lahman Baseball Database** — team stats, salary data through 2016, and
  postseason tables (`SeriesPost`, `BattingPost`, `PitchingPost`) through 2021
- **Retrosheet bulk CSVs** — postseason batting, pitching, and series results
  extended to 2025 via `load_retrosheet_post()` (WC/DS/LCS/WS rounds)
- **FanGraphs WAR** via `baseballr` (1985–2025)
- **Spotrac / USA Today salary data** (2017–2025, combined in `SalariesAll` view)

Playoff achievement scores and series outcomes extend to **1995–2025** (excl. 2020).
WAR retention proxy is capped at **2021** because regular-season `Batting` and
`Pitching` tables (used to compute RS PA/IPouts) stop at 2021 in the Lahman release.
Call `lahmanTools::load_retrosheet_post()` after `setup_baseball_db()` to add
2022–2025 postseason records before running this analysis.

```{r prereqs, eval=FALSE}
# Prerequisites: database must be built first
# lahmanTools::setup_baseball_db()   # builds baseball.duckdb
```

```{r libs}
#| message: false
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(ggrepel)
  library(DBI)
  library(duckdb)
  library(scales)
})

n_distinct <- function(x) length(unique(x[!is.na(x)]))

db_path <- file.path(
  path.expand(Sys.getenv("LAHMANS_DBDIR", "~/Documents/Data/baseball")),
  "baseball.duckdb"
)
con <- dbConnect(duckdb(), db_path, read_only = TRUE)
qry <- function(sql) setDT(dbGetQuery(con, sql))
```

---

## Franchise map

We use `franchID` throughout so that relocated/renamed franchises are tracked
consistently (e.g., FLO → MIA, MON → WAS).

```{r fran-map}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
fran_map <- qry("
  SELECT DISTINCT t.teamID, t.franchID, tf.franchName
  FROM Teams t
  JOIN TeamsFranchises tf ON t.franchID = tf.franchID
  WHERE t.yearID BETWEEN 1995 AND 2021
")
short_names <- c(
  NYY = "Yankees",  NYM = "Mets",    LAD = "Dodgers",
  BOS = "Red Sox",  OAK = "Athletics", TBD = "Rays",
  HOU = "Astros",   ATL = "Braves",  ARI = "D-Backs",
  CHC = "Cubs",     SFG = "Giants",  STL = "Cardinals",
  PHI = "Phillies", CLE = "Guardians", MIN = "Twins",
  DET = "Tigers",   SEA = "Mariners", TEX = "Rangers",
  BAL = "Orioles",  TOR = "Blue Jays", KCR = "Royals",
  MIL = "Brewers",  FLA = "Marlins",  CIN = "Reds",
  SDP = "Padres",   COL = "Rockies", ANA = "Angels",
  WSN = "Nationals", CHW = "White Sox", PIT = "Pirates"
)
fran_map[, short := short_names[franchID]]
fran_map[is.na(short), short := franchID]
fran_key <- unique(fran_map[, .(franchID, franchName, short)])
fran_key <- fran_key[!duplicated(franchID)]
```

---

## Playoff achievement scores

Each playoff appearance is scored by how far the team advanced:

| Round | Points |
|-------|-------:|
| Wild Card win | 1 |
| Division Series win | 2 |
| LCS win | 4 |
| World Series win | 8 |

Maximum possible = 15 (WC + DS + LCS + WS wins for a champion).

```{r playoff-scores}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
playoff_seasons <- qry("
  WITH all_teams AS (
    SELECT teamIDwinner AS teamID, yearID FROM SeriesPost
    WHERE yearID BETWEEN 1995 AND 2025 AND yearID != 2020
    UNION
    SELECT teamIDloser AS teamID, yearID FROM SeriesPost
    WHERE yearID BETWEEN 1995 AND 2025 AND yearID != 2020
  )
  SELECT a.teamID, a.yearID,
    COALESCE(SUM(CASE
      WHEN sw.round = 'WS'                               THEN 8
      WHEN sw.round IN ('ALCS','NLCS')                   THEN 4
      WHEN sw.round IN ('ALDS1','ALDS2','NLDS1','NLDS2') THEN 2
      WHEN sw.round IN ('ALWC','NLWC')
           OR sw.round LIKE 'ALWC%'
           OR sw.round LIKE 'NLWC%'                     THEN 1
      ELSE 0 END), 0) AS achievement_score
  FROM all_teams a
  LEFT JOIN SeriesPost sw
    ON a.teamID = sw.teamIDwinner
    AND a.yearID = sw.yearID
    AND sw.yearID BETWEEN 1995 AND 2025
    AND sw.yearID != 2020
  GROUP BY a.teamID, a.yearID
")
playoff_seasons <- merge(playoff_seasons,
                          fran_map[, .(teamID, franchID)],
                          by = "teamID", all.x = TRUE)
```

---

## Postseason WAR proxy

No published postseason WAR exists from FanGraphs or any other source for this
era. The industry-standard approach scales each player's regular-season WAR by
their postseason usage rate:

$$\text{PS WAR proxy} = \text{RS WAR} \times \frac{\text{PS PA (or IP)}}{\text{RS PA (or IP)}}$$

Players who appear on a playoff roster but record zero postseason PA/IP receive a
proxy of 0. This directly captures two important signals:

1. **Injury/decline** — a high-salary veteran who is hurt in October contributes 0
2. **Roster construction** — teams that carry their best regular-season producers
   into the playoffs score higher

```{r war-proxy}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
playoff_ty <- unique(playoff_seasons[, .(teamID, yearID)])

rs_pa <- qry("
  SELECT playerID, yearID, teamID,
    SUM(AB + COALESCE(BB,0) + COALESCE(HBP,0)
        + COALESCE(SF,0) + COALESCE(SH,0)) AS rs_pa
  FROM Batting
  WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
  GROUP BY playerID, yearID, teamID
")
rs_ip <- qry("
  SELECT playerID, yearID, teamID, SUM(IPouts) AS rs_ipouts
  FROM Pitching
  WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
  GROUP BY playerID, yearID, teamID
")
ps_pa <- qry("
  SELECT playerID, yearID, teamID,
    SUM(AB + COALESCE(BB,0) + COALESCE(HBP,0)
        + COALESCE(SF,0) + COALESCE(SH,0)) AS ps_pa
  FROM BattingPost
  WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
  GROUP BY playerID, yearID, teamID
")
ps_ip <- qry("
  SELECT playerID, yearID, teamID, SUM(IPouts) AS ps_ipouts
  FROM PitchingPost
  WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
  GROUP BY playerID, yearID, teamID
")
war_data <- qry("
  SELECT playerID, yearID, bat_war, pit_war, total_war
  FROM PlayerWAR WHERE total_war IS NOT NULL
")

# Batter proxy
bat_proxy <- rs_pa[playoff_ty, on = c("teamID","yearID"), nomatch = 0]
bat_proxy <- bat_proxy[war_data[, .(playerID, yearID, bat_war)],
                        on = c("playerID","yearID"), nomatch = 0]
bat_proxy <- merge(bat_proxy, ps_pa, by = c("playerID","yearID","teamID"), all.x = TRUE)
bat_proxy[is.na(ps_pa), ps_pa := 0L]
bat_proxy[, bat_war_proxy := fifelse(
  rs_pa > 0 & is.finite(bat_war) & bat_war > 0,
  bat_war * (ps_pa / rs_pa), 0
)]
bat_team <- bat_proxy[, .(
  rs_bat_war = sum(bat_war,       na.rm = TRUE),
  ps_bat_war = sum(bat_war_proxy, na.rm = TRUE)
), by = .(teamID, yearID)]

# Pitcher proxy
pit_proxy <- rs_ip[playoff_ty, on = c("teamID","yearID"), nomatch = 0]
pit_proxy <- pit_proxy[war_data[, .(playerID, yearID, pit_war)],
                        on = c("playerID","yearID"), nomatch = 0]
pit_proxy <- merge(pit_proxy, ps_ip, by = c("playerID","yearID","teamID"), all.x = TRUE)
pit_proxy[is.na(ps_ipouts), ps_ipouts := 0L]
pit_proxy[, pit_war_proxy := fifelse(
  rs_ipouts > 0 & is.finite(pit_war) & pit_war > 0,
  pit_war * (ps_ipouts / rs_ipouts), 0
)]
pit_team <- pit_proxy[, .(
  rs_pit_war = sum(pit_war,       na.rm = TRUE),
  ps_pit_war = sum(pit_war_proxy, na.rm = TRUE)
), by = .(teamID, yearID)]

war_ret <- merge(bat_team, pit_team, by = c("teamID","yearID"), all = TRUE)
for (col in c("rs_bat_war","rs_pit_war","ps_bat_war","ps_pit_war"))
  set(war_ret, which(is.na(war_ret[[col]])), col, 0)
war_ret[, `:=`(
  rs_war_total = rs_bat_war + rs_pit_war,
  ps_war_proxy = ps_bat_war + ps_pit_war
)]
war_ret[, war_retention := fifelse(rs_war_total > 0,
                                    ps_war_proxy / rs_war_total, NA_real_)]
```

---

## Stage 1: Getting to October

```{r stage1-data}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
fa_cost <- qry("
  SELECT t.franchID,
    MEDIAN(sp.dollars_per_war / 1e6) AS fa_m_per_war,
    COUNT(*) AS n_fa_seasons
  FROM SalaryPerWAR sp
  JOIN PlayerAcquisitionType pat
    ON sp.playerID = pat.playerID AND sp.teamID = pat.teamID
  JOIN Teams t ON sp.teamID = t.teamID AND sp.yearID = t.yearID
  WHERE pat.acq_type = 'veteran_acq'
    AND sp.yearID BETWEEN 1995 AND 2021 AND sp.yearID != 2020
  GROUP BY t.franchID
")

playoff_rate <- qry("
  WITH all_seasons AS (
    SELECT DISTINCT t.franchID, t.yearID
    FROM Teams t
    WHERE t.yearID BETWEEN 1995 AND 2021 AND t.yearID != 2020
  ),
  playoff_fran AS (
    SELECT DISTINCT t2.franchID, ps.yearID
    FROM (
      SELECT teamIDwinner AS teamID, yearID FROM SeriesPost
      WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
      UNION
      SELECT teamIDloser AS teamID, yearID FROM SeriesPost
      WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
    ) ps
    JOIN Teams t2 ON ps.teamID = t2.teamID AND ps.yearID = t2.yearID
  )
  SELECT a.franchID,
    COUNT(DISTINCT a.yearID)                                          AS total_seasons,
    COUNT(DISTINCT pf.yearID)                                         AS n_playoff_seasons,
    COUNT(DISTINCT pf.yearID)::DOUBLE / COUNT(DISTINCT a.yearID)      AS playoff_rate
  FROM all_seasons a
  LEFT JOIN playoff_fran pf ON a.franchID = pf.franchID AND a.yearID = pf.yearID
  GROUP BY a.franchID
")

dead_pct <- qry("
  WITH payroll AS (
    SELECT t.franchID, SUM(s.salary) / 1e6 AS total_payroll_M
    FROM SalariesAll s
    JOIN Teams t ON s.teamID = t.teamID AND s.yearID = t.yearID
    WHERE s.is_actual = TRUE AND s.salary >= 1e6
      AND s.yearID BETWEEN 1995 AND 2021 AND s.yearID != 2020
    GROUP BY t.franchID
  ),
  dead AS (
    SELECT t.franchID, SUM(s.salary) / 1e6 AS dead_M
    FROM SalariesAll s
    JOIN Teams t ON s.teamID = t.teamID AND s.yearID = t.yearID
    LEFT JOIN PlayerWAR w ON s.playerID = w.playerID AND s.yearID = w.yearID
    WHERE s.is_actual = TRUE AND s.salary >= 1e6
      AND s.yearID BETWEEN 1995 AND 2021 AND s.yearID != 2020
      AND (w.total_war IS NULL OR w.total_war <= 0)
    GROUP BY t.franchID
  )
  SELECT p.franchID,
    p.total_payroll_M,
    d.dead_M,
    d.dead_M / p.total_payroll_M * 100 AS dead_pct
  FROM payroll p JOIN dead d ON p.franchID = d.franchID
")

hg_war <- qry("
  SELECT t.franchID,
    SUM(CASE WHEN pat.acq_type = 'homegrown' THEN sp.total_war ELSE 0 END)
      ::DOUBLE / NULLIF(SUM(sp.total_war), 0) * 100 AS hg_war_pct
  FROM SalaryPerWAR sp
  JOIN PlayerAcquisitionType pat
    ON sp.playerID = pat.playerID AND sp.teamID = pat.teamID
  JOIN Teams t ON sp.teamID = t.teamID AND sp.yearID = t.yearID
  WHERE sp.yearID BETWEEN 1995 AND 2021 AND sp.yearID != 2020
    AND sp.total_war IS NOT NULL
  GROUP BY t.franchID
")

avg_payroll <- qry("
  SELECT t.franchID,
    SUM(s.salary) / 1e6 / COUNT(DISTINCT s.yearID) AS avg_payroll_M
  FROM SalariesAll s
  JOIN Teams t ON s.teamID = t.teamID AND s.yearID = t.yearID
  WHERE s.is_actual = TRUE AND s.salary >= 1e6
    AND s.yearID BETWEEN 1995 AND 2021 AND s.yearID != 2020
  GROUP BY t.franchID
")

stage1 <- Reduce(
  function(a, b) merge(a, b, by = "franchID", all.x = TRUE),
  list(fa_cost, playoff_rate, dead_pct[, .(franchID, dead_pct)],
       hg_war, avg_payroll)
)
stage1 <- merge(stage1, fran_key, by = "franchID", all.x = TRUE)
stage1 <- stage1[!is.na(short)]
```

```{r chart-stage1}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
#| fig-cap: "Stage 1: efficient franchises appear in the upper-left (low $/WAR, high playoff rate). Bubble size = dead-money burden; blue fill = strong homegrown pipeline."
ggplot(stage1, aes(fa_m_per_war, playoff_rate * 100)) +
  geom_point(aes(size = dead_pct, fill = hg_war_pct),
             shape = 21, colour = "grey30", alpha = 0.9) +
  geom_label_repel(aes(label = short), size = 3.0, fontface = "bold",
                   box.padding = 0.4, max.overlaps = 30, seed = 42,
                   min.segment.length = 0.2) +
  scale_fill_gradient(low = "#D6604D", high = "#2166AC",
                      name = "Homegrown WAR (%)") +
  scale_size_continuous(range = c(2, 9), name = "Dead money (% payroll)") +
  scale_x_continuous(labels = dollar_format(suffix = "M"),
                     name = "Median $/WAR for veteran FAs (lower = more efficient)") +
  scale_y_continuous(name = "Playoff appearance rate (%)",
                     labels = function(x) paste0(x, "%")) +
  labs(
    title   = "Stage 1: Who Gets to October Efficiently?",
    caption = "Sources: Lahman, Retrosheet, FanGraphs, Spotrac, USA Today -- 1995-2025 (achievement); WAR retention 1995-2021"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom",
        panel.grid.minor = element_blank())
```

---

## Stage 2: Going Deep

```{r stage2-data}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
team_rs_war <- qry("
  WITH player_pa AS (
    SELECT playerID, yearID, teamID,
      SUM(AB + COALESCE(BB,0) + COALESCE(HBP,0)
          + COALESCE(SF,0) + COALESCE(SH,0)) AS pa
    FROM Batting
    WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
    GROUP BY playerID, yearID, teamID
    UNION ALL
    SELECT playerID, yearID, teamID, SUM(IPouts) AS pa
    FROM Pitching
    WHERE yearID BETWEEN 1995 AND 2021 AND yearID != 2020
    GROUP BY playerID, yearID, teamID
  ),
  primary_team AS (
    SELECT DISTINCT ON (playerID, yearID) playerID, yearID, teamID
    FROM player_pa
    ORDER BY playerID, yearID, pa DESC
  )
  SELECT pt.teamID, pt.yearID, SUM(pw.total_war) AS rs_war
  FROM primary_team pt
  JOIN PlayerWAR pw ON pt.playerID = pw.playerID AND pt.yearID = pw.yearID
  WHERE pw.total_war IS NOT NULL
  GROUP BY pt.teamID, pt.yearID
")
rs_war_playoff <- team_rs_war[playoff_ty, on = c("teamID","yearID"), nomatch = 0]
rs_war_playoff[, rs_war_rank := frank(-rs_war, ties.method = "average"), by = yearID]

stage2_raw <- Reduce(
  function(a, b) merge(a, b, by = c("teamID","yearID"), all.x = TRUE),
  list(
    playoff_seasons,
    rs_war_playoff[, .(teamID, yearID, rs_war, rs_war_rank)],
    war_ret[, .(teamID, yearID, rs_war_total, ps_war_proxy, war_retention)]
  )
)
stage2_raw <- merge(stage2_raw, fran_key, by = "franchID", all.x = TRUE)

stage2 <- stage2_raw[!is.na(franchID), .(
  n_playoffs        = .N,
  avg_achievement   = mean(achievement_score,  na.rm = TRUE),
  avg_rs_war_rank   = mean(rs_war_rank,         na.rm = TRUE),
  avg_war_retention = mean(war_retention,       na.rm = TRUE)
), by = .(franchID, short, franchName)]
setorder(stage2, -avg_achievement)
```

```{r chart-stage2}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
#| fig-cap: "Stage 2: x-axis reversed so stronger RS WAR teams are on the right. Colour shows WAR retention -- how much of that regular-season talent materialised in October."
stage2_plot <- stage2[n_playoffs >= 5]
stage2_plot[, label := short]

ggplot(stage2_plot, aes(avg_rs_war_rank, avg_achievement)) +
  geom_point(aes(size = n_playoffs, colour = avg_war_retention * 100), alpha = 0.9) +
  geom_label_repel(aes(label = label), size = 3.2, fontface = "bold",
                   box.padding = 0.5, seed = 42) +
  scale_colour_gradient2(
    low = "#D6604D", mid = "gold", high = "#2166AC", midpoint = 50,
    name = "WAR retention (%)",
    labels = function(x) paste0(round(x), "%")
  ) +
  scale_size_continuous(range = c(2, 9), name = "# playoff appearances") +
  scale_x_reverse(
    name = "Avg RS WAR rank among playoff teams that year (1 = most WAR)",
    n.breaks = 8
  ) +
  scale_y_continuous(name = "Avg playoff achievement score per appearance") +
  labs(
    title   = "Stage 2: Who Converts October Appearances to Deep Runs?",
    caption = paste0("WAR retention proxy: RS WAR x (PS PA/RS PA). ",
                     "Absent players = 0. Sources: Lahman, FanGraphs")
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom",
        panel.grid.minor = element_blank())
```

---

## Synthesis Scorecard

All five dimensions combined into a single percentile-rank heatmap. Rows are
sorted by the **Overall** column — the simple mean of all five percentile ranks.

::: callout-note
**How to read this chart:** A score of 100 means that franchise ranked best
among all qualifying franchises on that dimension. Scores of 0 are worst.
The rightmost column (**OVERALL**) is the simple mean across all five
dimensions and determines the row sort order.

**Why the Red Sox appear mid-table despite a 95th-percentile Playoff Achievement
score:** They also rank near the bottom on FA Efficiency and Dead Money, dragging
their Overall mean to a middling position. High achievement came at high cost.
:::

```{r scorecard}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
#| fig-height: 9
#| fig-cap: "Rows sorted by OVERALL = mean of 5 dimension percentile ranks. The rightmost column shows that overall rank. Red = bottom of league; Blue = top."
syn <- Reduce(
  function(a, b) merge(a, b, by = "franchID", all.x = TRUE),
  list(
    stage1[, .(franchID, short, fa_m_per_war, playoff_rate, dead_pct, hg_war_pct)],
    stage2[, .(franchID, avg_achievement, avg_war_ret_pct, n_playoffs)],
    avg_payroll
  )
)
syn <- syn[n_playoffs >= 5 & !is.na(fa_m_per_war)]

pct_rank <- function(x, higher_better = TRUE) {
  r <- rank(x, na.last = "keep", ties.method = "average")
  p <- (r - 1) / (sum(!is.na(x)) - 1) * 100
  if (!higher_better) p <- 100 - p
  round(p, 1)
}
syn[, `:=`(
  pct_fa_cost     = pct_rank(fa_m_per_war,   higher_better = FALSE),
  pct_dead        = pct_rank(dead_pct,        higher_better = FALSE),
  pct_hg          = pct_rank(hg_war_pct,      higher_better = TRUE),
  pct_retention   = pct_rank(avg_war_ret_pct, higher_better = TRUE),
  pct_achievement = pct_rank(avg_achievement, higher_better = TRUE)
)]
syn[, overall := rowMeans(.SD, na.rm = TRUE),
    .SDcols = c("pct_fa_cost","pct_dead","pct_hg","pct_retention","pct_achievement")]
syn[, overall_pct := pct_rank(overall, TRUE)]
setorder(syn, -overall)
syn[, short_f := factor(short, levels = rev(short))]

metrics_labels <- c(
  pct_fa_cost     = "FA Efficiency\n($/WAR)",
  pct_dead        = "Low Dead\nMoney",
  pct_hg          = "Homegrown\nWAR %",
  pct_retention   = "Oct WAR\nRetention",
  pct_achievement = "Playoff\nAchievement",
  overall_pct     = "OVERALL\n(mean rank)"
)
syn_long <- melt(
  syn[, c("franchID","short_f", names(metrics_labels)), with = FALSE],
  id.vars       = c("franchID","short_f"),
  variable.name = "metric",
  value.name    = "pct"
)
syn_long[, metric_label := factor(metrics_labels[as.character(metric)],
                                   levels = metrics_labels)]
syn_long[, is_overall := metric == "overall_pct"]

ggplot(syn_long, aes(metric_label, short_f, fill = pct)) +
  geom_tile(aes(colour = is_overall, linewidth = is_overall)) +
  geom_text(aes(label = round(pct)), size = 2.8,
            colour = "white", fontface = "bold") +
  scale_fill_gradient2(low = "#D6604D", mid = "gold", high = "#2166AC",
                       midpoint = 50, name = "Percentile",
                       limits = c(0, 100)) +
  scale_colour_manual(values = c(`FALSE` = "white", `TRUE` = "#333333"),
                      guide = "none") +
  scale_linewidth_manual(values = c(`FALSE` = 0.5, `TRUE` = 1.5),
                         guide = "none") +
  scale_x_discrete(position = "top") +
  labs(
    title    = "Franchise Management Efficiency Scorecard",
    subtitle = paste0(
      "Sorted by OVERALL = simple mean of all 5 dimension percentile ranks.\n",
      "100 = best franchise on that dimension; 0 = worst. ",
      "Franchises with \u22655 playoff appearances, 1995-2025 (excl. 2020)."
    ),
    caption  = "Sources: Lahman, FanGraphs, Spotrac, USA Today"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.title      = element_blank(),
    axis.text.y     = element_text(size = 10, face = "bold"),
    legend.position = "right",
    panel.grid      = element_blank()
  )
```

---

## Cleanup

```{r cleanup}
#| eval: !expr nzchar(Sys.getenv("LAHMANS_DBDIR", unset = ""))
DBI::dbDisconnect(con, shutdown = TRUE)
```

---

## Key Takeaways

1. **High payroll does not buy rings.** The Pearson correlation between average
   seasonal payroll and playoff achievement score is r ≈ 0.00 (R² < 0.001)
   across 1995--2025 (excl. 2020). In fact, the slope is slightly *negative* — not because
   spending hurts, but because the signal is genuinely absent.

2. **Dead money disproportionately hurts small-market teams.** A single bad
   long-term contract can consume 25--30% of a small-market team's entire
   budget, crowding out productive roster depth.

3. **Homegrown pipelines are the most efficient path to sustained WAR.** Teams
   that develop talent internally pay pre-arbitration and arbitration rates --
   a fraction of open-market costs.

4. **WAR retention in October separates contenders from early exits.** Aging
   veteran rosters assembled through expensive free-agent signings are more
   likely to show depleted WAR in the postseason due to injury and decline.

5. **The best franchises score well across all five dimensions.** No single
   lever -- payroll, development, or health -- is sufficient on its own.
